Two subtle problems with over-representation analysis

### 2024-07-09 manuscript/manuscript.Rmd
ms

(*) Corresponding author: mark.ziemann@burnet.edu.au

@@

The background problem came to our attention when a doctoral student in our department
was puzzled by results obtained from clusterProfiler when using a custom set of pathways.
This exact issue was the subject of several posts to online bioinformatics forums dating to
~2020 where confused users were wondering why the number of background genes changed so much.
The results shown here show that the background problem affects smaller gene set libraries
like KEGG and HPO more severely as compared to Cellmarkers and GO libraries which describe
a larger proportion of all genes.
ClusterProfiler maintainers recently provided an option to prevent the loss of unannotated
genes from the background `options(enrichment_force_universe = TRUE)`, however this feature
isn't yet described in the official documentation.

@@

@@

The purpose of this work was to describe two subtle problems with ORA, but there are some
bigger problems noted.
Firstly, all variations of ORA were less accurate as compared to FCS in this simulation work, confirming a
previous report [@Kaspi2020-rn].
Specifically, precision of ORA deteriorated with greater levels of added noise, while FCS
was relatively stable.
This means researchers can be more confident with FCS results, even if the data are
noisy.
Higher recall of FCS also means that researchers will have more power to identify pathways
relevant to the biological processes being studied.
Secondly, many ORA tools do not provide enrichment scores in their default output tables.
Enrichment scores are a surrogate for effect size, and without them, users may over-rely
on significance scores when interpreting enrichment results [@Ziemann2024-ts].
Lastly, ORA results can vary dramatically depending on the number of genes selected in the
foreground, a mostly arbitrary choice.

@@

## Competing Interests

@@

## Funding

@@

## Author Contributions

@@

## Acknowledgements

>>>>

(*) Corresponding author: mark.ziemann@burnet.edu.au Burnet Institute, 85 Commercial Rd Melbourne VIC 3004, Australia.

@@

The background problem came to our attention when a doctoral student in our department
was puzzled by results obtained from clusterProfiler when using a custom set of pathways.
This exact issue was the subject of several posts to online bioinformatics forums dating to
~2020 where confused users were wondering why the number of background genes changed so much.
The results shown here indicate the background problem affects smaller gene set libraries
like KEGG and HPO more severely as compared to Cellmarkers and GO libraries which describe
a larger proportion of all genes.
ClusterProfiler maintainers recently provided an option to prevent the loss of unannotated
genes from the background `options(enrichment_force_universe = TRUE)`, however this feature
isn't yet described in the official documentation.

@@

So end-users can explore the effect of these two problems on their work, we have developed
an R/Shiny web application.
Users can upload their foreground and background lists, select their preferred gene set
library and compare results from default and patched analyses.
The resource is currently available at https://oratool.ziemann-lab.net/ and available as a
docker container (https://hub.docker.com/repository/docker/mziemann/background_app/general).

@@

The purpose of this work was to thoroughly characterise two subtle problems with ORA, but there are some
bigger problems noted.
Firstly, all variations of ORA were less accurate as compared to FCS in this simulation work, confirming a
previous report [@Kaspi2020-rn].
Specifically, precision of ORA deteriorated with greater levels of added noise, while FCS
was relatively stable.
This means researchers can be more confident with FCS results, even if the data are
noisy.
Higher recall of FCS also means that researchers will have more power to identify pathways
relevant to the biological processes being studied.
Secondly, many ORA tools do not provide enrichment scores in their default output tables.
Enrichment scores are a surrogate for effect size, and without them, users may over-rely
on significance scores when interpreting enrichment results [@Ziemann2024-ts].
Lastly, ORA results can vary dramatically depending on the number of genes selected in the
foreground, a mostly arbitrary choice.

@@

### Competing Interests

@@

### Funding

@@

### Author Contributions

@@

### Acknowledgements
